---
title: "Project 3 - Linear Regression"
author: "Carlos Ribadeneira"
date: ""
output:
  html_document:
    df_print: paged
---
## Question 1: Theoretical Questions about OLS

 In this example below, x is how much you `spend` on advertisement and y is the number of `sales` generated. For each part below, please, show your code.

```{r echo=TRUE, eval=TRUE}
sales <- c(9914,40487,54324,50044,34719,42551,94871,118914,158484,131348,78504,36284)
spend <- c(1000,4000,5000,4500,3000,4000,9000,11000,15000,12000,7000,3000)

mydata<-data.frame(sales,spend)
```

- Calculate average of your X variable and save the variable as `mean_spend`. Report the mean of X.


```{r}
mean_spend<-mean(mydata$spend)
```
The mean of the x variable is `r mean(mydata$spend)`.

- Calculate the difference between each X and the average X and save it as a new column in your data called `dev_x`.

```{r}
mydata$dev_x<-mydata$spend-mean_spend
```
The difference of the x variable and the average is `r mydata$spend-mean_spend`.

- Square the differences (i.e. square each row of `dev_x`) and add it all up and save as a new variable called SSxx. (SS stands for Sum of Squares). Report SSxx.

```{r}
SSxx<- sum(mydata$dev_x^2)
```
The SSxx of the  variable is `r sum(mydata$dev_x^2)`.

- Calculate average of your Y variable and save the variable as `mean_sales`. Report the mean of Y.

```{r}
mean_sales<-mean(mydata$sales)
```
The mean of Y is `r mean(mydata$sales)`.

- Calculate the difference between each Y and the average Y and save it as a new column in your data called `dev_y`.

```{r}
mydata$dev_y<-mydata$sales-mean_sales
```
The difference between each Y is `r mydata$sales-mean_sales`.

- Multiply the differences (`dev_y` * `dev_x`) and add them all together.  Call this new variable SSxy.

```{r}
SSxy<-sum(mydata$dev_y*mydata$dev_x)
```
SSxy is `r sum(mydata$dev_y*mydata$dev_x)`.

- Calculate the regression slope as SSxy / SSxx. Report the slope

```{r}
slope<-SSxy/SSxx
```
Slope is `r SSxy/SSxx`.

- Calculate the intercept by subtracting (SSxy / SSxx) * AVG(X) from AVG(Y). Report the intercept

```{r}
intercept<-mean_sales-slope*mean_spend
```
The intercept is `r mean_sales-slope*mean_spend`. 

- Using the regression intercept and slope you have found, create a new column called `y_hat` which will include the predicted values of y given the values of x in each row.

```{r}
mydata$y_hat<-intercept+slope*mydata$spend
```
y_hat is `r intercept+slope*mydata$spend`.

- Calculate a new column called resid_1 and calculate the difference between `sales` - `y_hat`.

```{r}
mydata$resid_1<-mydata$sales-mydata$y_hat
```
Difference between sales and y_hat is `r mydata$sales-mydata$y_hat`.

- Let's find the variance of the error term. The resid_1 column contains the residuals of the regression. We can calculate the variance by first squaring each residual, next summing up all of the values, and finally dividing by the number of observations minus the number of betas estimated, which is two in this case (constant and slope). Save the variance as var_resid. Report the square root of var_resid.

```{r}
var_resid<-sum(mydata$resid_1^2)/(nrow(mydata)-2)
sqrt(var_resid)
```
var_resid is `r sqrt(var_resid)`.

- Next, let's find the $R^2$. The residuals tell us what the regression line doesn't explain. Calculate the sum of squares for the `resid_1` column and save as SSresid. Calculate the sum of squares for the `sales` column. Call this value SSyy. We can then calculate $R^2$ as 1- SSresid/SSyy. Report the $R^2$.

```{r}
SSresid<-sum(mydata$resid_1^2)
SSyy<-sum(mydata$dev_y^2)
R2<- 1-SSresid/SSyy
```
R_squared is `r 1-SSresid/SSyy`.

- Now let's calculate the coefficients and standard error using matrix operations.
```{r echo=TRUE, eval=FALSE}
# We need to include a column of ones for the constant
X<-matrix(c(rep(1,length(sales)),spend),12,2)
# This is equivalent to COV(X,Y)/VAR(X)
beta<-solve(t(X)%*%X)%*%t(X)%*%sales
# These are the predicted values of Y
y_hat <- X%*%beta
# These are the calculated residuals
resid_2 <- sales - y_hat
# We are calculating the variance of the residuals
var_reg <-(t(resid_2)%*%resid_2)/10
# Standard errors of the coefficients
b_se <- sqrt(diag(solve(t(X)%*%X)*var_reg[1,1]))
# check your work using lm
summary(lm(sales~spend))
```
You should check your work for question 1. Everyone should get this first question completely correct.

## Question 2
In this question, you will use linear regression to study speeding tickets. Each question builds on the previous question. Your regressions should have more controls as you move through the assignment. Try to capture all of these regression in one nicely formatted table.

What determines how much drivers are fined if they are stopped for speeding? Do demographics like age, gender, and race matter? To answer this question, we'll investigate traffic stops and citations in Massachusetts using data from Makowsky and Stratmann (2009). Even though state law sets a formula for tickets based on how fast a person was driving, police officers in practice often deviate from the formula. An amount for the fine is given only for observations in which the police officer decided to assess a fine.


```{r echo=FALSE}
library(stargazer)
library(dplyr)
library(ggplot2)
library(haven)
library(car)
speed<- read.csv("speeding_tickets_text.csv")

```

a) Plot a histogram of fines. Does it looked normally distributed or skewed? Keep in mind to remove missing fine amounts.
```{r}
speed <- speed %>%na.omit()
hist(speed$Amount, xlab = "Fine amount $", main = "Speeding Fines")
```
The distribution is skewed to right. In other words, the mode and median are less than the mean. The distribution shows that there are outliers in the right tail. 

b) Estimate a simple linear regression model in which the ticket amount is the dependent variable as a function of age. Is age statistically significant? 
```{r results='asis'}
reg1<-lm(Amount~Age, data=speed)
stargazer(reg1, type='html')
```
Age is statistically significant in this particular model. 

c) What does it mean for a variable to be endogenous? Is it possible the variable "age" is endogenous? Please explain your answer.

Endogenous variables have values that are determined by one or more of the independent variables in the system or model. 
It is a very good possibility that "age" is endogenous. There are other factors in this model that can possibly be related to age. As age changes it can create a change in another independent variable, therefore making them endogenous. For example, MPHover can have a relationship with age such that younger individuals tend to drive faster than older individuals. 


d) Are miles per hour over the speed limit correlated with age? Report the correlation coefficient.
```{r}
cor1 <- cor(speed$MPHover,speed$Age)
```
Miles per hour over the speed limit and age are correlated with a value of 'r cor1'.


e) Estimate the model from part b), also controlling for miles per hour over the speed limit. Explain what happens to the coefficient on age and why. 
```{r results='asis'}
reg2<-lm(Amount~Age+MPHover, data = speed)
stargazer(reg1, reg2, type='html')
```
The coefficient of age is no longer significant once the MPH over variable is included in the model. This is because age could have been encompassing the data of MPH over when Age was the only variable in the model.


f) Is the effect of age on fines linear or non-linear? Assess this question by estimating a model that includes both age and a quadratic age term. Also, control for _MPHover, Female, Black, and Hispanic_. Interpret the coefficients on the age variables. Remember it is non-linear so you can't just read the coefficient.
```{r results='asis'}
speed$Age2<- speed$Age^2
reg3<-lm(Amount~Age2+Age+MPHover+Black+Hispanic+Female, data = speed)
stargazer(reg1,reg2,reg3,type='html')
```
The effect of age on fines is non-linear. The coefficient of the quadratic age term is positive, meaning that the curve is concave (upwards facing). 


g) Plot the relationship between age and ticket amount using the coefficients on age that you found in step f. Calculate the fitted value for a white male with 0 _MPHover_ (probably not many people going zero miles over the speed limit got a ticket, but this simplifies calculations a lot) for ages equal to 20, 25, 30, 35, 40, and 70. Use R to calculate these values and plot them. 
plug and chug
```{r}
Age1<- 20
Age2<- 25
Age3<- 30
Age4<- 35
Age5<- 40
Age6<- 70

Amount_Age20 <- 9.867 + 0.004*(Age1^2) -0.265*(Age1) + 6.873*(0) - 1.920*(0) + 2.068*(0) - 3.510*(0)
Amount_Age25 <- 9.867 + 0.004*(Age2^2) -0.265*(Age2) + 6.873*(0) - 1.920*(0) + 2.068*(0) - 3.510*(0)
Amount_Age30 <- 9.867 + 0.004*(Age3^2) -0.265*(Age3) + 6.873*(0) - 1.920*(0) + 2.068*(0) - 3.510*(0)
Amount_Age35 <- 9.867 + 0.004*(Age4^2) -0.265*(Age4) + 6.873*(0) - 1.920*(0) + 2.068*(0) - 3.510*(0)
Amount_Age40 <- 9.867 + 0.004*(Age5^2) -0.265*(Age5) + 6.873*(0) - 1.920*(0) + 2.068*(0) - 3.510*(0)
Amount_Age70 <- 9.867 + 0.004*(Age6^2) -0.265*(Age6) + 6.873*(0) - 1.920*(0) + 2.068*(0) - 3.510*(0)

Age_Tickets <- data.frame("Age" = c(Age1, Age2, Age3, Age4, Age5, Age6), "Amount" = c(Amount_Age20, Amount_Age25, Amount_Age30, Amount_Age35, Amount_Age40, Amount_Age70))

Age_Tickets_plot <- ggplot(Age_Tickets, aes(x = Age, y = Amount))+ggtitle("Age vs Fine Amount") + geom_point()+ geom_smooth(method = "lm")

Age_Tickets_plot
```
It is worthwhile noting in the plot below that using the criteria previouly mentioned, as age increases, the amount decreases. However, after age 30 the ticket amount starts to increase. As confirmed, in question f the distribution shows a non-linear effect. 

h) Calculate the age that is associated with the lowest predicted fines. __Hint: You can use calculus or an algebraic formula used to find the minimum and maximum of quadratic functions.__ $$y =ax^2 + bx +c \\ x = -\frac{b}{2a}$$
```{r}
b <- -0.264765
a <- 0.003743

x = (-b)/(2*a)
x

```
The age associated with the lowest predicted fines is 'r x'


i) Do drivers from out of town  and out of state get treated differently? Do state police and local police treat non-locals differently? Estimate a model that allows us to assess whether out of towners and out of staters are treated differently and whether state police respond differently to out of towners and out of staters. Interpret the coefficients on the relevant variables. __Hint: you have to do something more than just including the dummy variables.__
local police and intowners left out
```{r results='asis'}
reg6<-lm(Amount~Age2+Age+MPHover+Black+Hispanic+Female+StatePol, data = speed)
reg7<-(lm(Amount~Age2+Age+MPHover+Black+Hispanic+Female+OutTown+OutState, data = speed))
reg8<-(lm(Amount~Age2+Age+MPHover+Black+Hispanic+Female+OutTown+OutState+StatePol, data = speed))
reg4<-(lm(Amount~Age2+Age+MPHover+Black+Hispanic+Female+OutTown+OutState+StatePol+OutTown*StatePol+OutState*StatePol, data = speed))
stargazer(reg4,reg6,reg7,reg8, type = "html")

```

This model reveals several different insights about the amount individuals are fined for speeding. First, the model reveals that individuals who are pulled over that are from out of state or out of town are treated differently. The coefficients from the 3rd regression model (Amount (3)-reg7) shows that out-of-towners get charged on average 4.379 dollars more than the locals, while the out-of-staters get charged on average 9.097 dollars more than the locals. This results in out-of-staters getting charged ~$5 more than out-of-towners. 

Secondly, the state police do treat nonlocals differently than the local police force because the state police tend to fine a higher amount to out-of-towners rather than out-of-staters, but the local police force tends to fine a higher amount to out-of-staters rather than out-of-towners. This is evident in the fact that the state police and out-of-towner interaction term has a coefficient of 6.173 (reference Amount #1-reg) but the state police and out of state interaction term is not significant (reference Amount (1)-reg). However, the local police have a significant coefficient of 7.7888 (reference Amount (4)-reg8)  for out-of-staters, but a low coefficient of 2.513 (reference Amount (4)-reg8) for out-of-towners.Therefore, the local police is charging out-of-staters more than out-of-towners and more than state police is charging any non-locals.


j) Test whether the two state police interaction terms are jointly significant. Briefly explain your results. __Hint: it says `jointly` so it is not a T-test.__
```{r}
reg5<-(lm(Amount~Age2+Age+MPHover+Black+Hispanic+Female+OutTown+OutState+StatePol+OutTown*StatePol+OutState*StatePol, data = speed))
linearHypothesis(reg5, c("OutTown:StatePol=0","OutState:StatePol=0"))
```


\newpage

Variable Name | Description
--------------|-----------------------
MPHover       | Miles per hour over the speed limit
Amount        | Assessed fine for the ticket
Age | Age of driver
Female | Equals 1 for women and 0 for men
Black | Equals 1 for African-American and 0 otherwise
Hispanic | Equals 1 for Hispanics and 0 otherwise
State Pol | Equals 1 if ticketing officer was state patrol officer and 0 otherwise
OutTown | Equals 1 if driver from out of town and 0 otherwise
OutState | Equals 1 if driver from out of state and 0 otherwise